---
title: "Bank Customer Churn Analysis Report"
author: "Maria Aslam"
date: "`r Sys.Date()`"
output:
html_document:
    toc: true
    toc_depth: 2
    theme: flatly
---



# Executive Summary 

### Strategic Overview 
This report analyzes customer attrition across our European banking operations. While France and Spain maintain stable churn rates (~16%), the bank faces a critical emergency in the German market, where churn has spiked to 32.4%. This analysis shifts focus from *who is leaving* to *why our most valuable segments are exiting*. 

### Key Findings 
* **The Two-Product Anchor:** We have identified a Survival Sweet Spot where customers with exactly 2 products have the highest loyalty (11–13% churn). 
* **Product Over-Saturation:** A catastrophic failure occurs when German customers reach 3 or 4 products, resulting in a 90–100% exit rate. This suggests our complex bundles are driving customers away rather than locking them in. 
* **Premium Segment Flight:** Churn is not limited to low-value accounts. High-balance customers and those in the "Very Good" credit tier (~33.5% churn) are leaving at nearly the same rate as high-risk segments. 
* **Service Failure:** There is a near-total correlation between complaints and attrition. Currently, a complaint is a "guaranteed exit signal," indicating our resolution process is failing to recover dissatisfied users.

### Core Recommendations 
* **Plus-One Retention:** Focus marketing on moving 1-product holders to 2 products; avoid pushing to 3+ until the *Product Fatigue* issue is resolved. 
* **German Priority Desk:** Launch a localized retention program for high-credit/high-balance German users to counter aggressive competitor poaching. 
* **Active Recovery:** Implement a proactive "Re-engagement" protocol for inactive German users before they hit the 41.1% churn threshold just check executive sumary section? is ti ok just excutive summary focus


# Data Cleaning & Methodology

To ensure the integrity of this analysis, the raw dataset underwent a rigorous preprocessing phase within the R environment. This ensures that all statistical summaries and visualizations reflect accurate customer behavior.

### The Cleaning Process
* **Dimensionality Reduction:** Irrelevant features such as `RowNumber`, `CustomerId`, and `Surname` were removed. These variables contain unique identifiers that do not contribute to churn patterns and would introduce "noise" into the analysis.
* **Attribute Re-classification:** Several columns were imported as integers but represent categorical states. I converted Exited, HasCrCard, IsActiveMember, and Complain into Factors. This allows the analysis to treat them as distinct groups rather than continuous numerical values.
* **Data Consistency Check:** I verified the Geography column for missing values and spelling inconsistencies to ensure the reliability of the regional "Deep Dive" results.
* **Metric Calculation Logic:** During the summarization of churn rates, a nested `as.numeric(as.character())` transformation was used. This ensures that binary factor levels are correctly interpreted for accurate percentage calculations.



```{r setup, include=FALSE}
library("tidyverse")
library("ggplot2")
library("knitr")
library("dplyr")
library("tidyr")
library("readr")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r, include=FALSE}
churn_rec <- read_csv("Customer-Churn-Records.csv")

```

```{r, include=FALSE}
bank_churn_cleaned <- churn_rec %>%
  select(-RowNumber, -CustomerId, -Surname) %>%
  mutate(across(c(Exited, HasCrCard, IsActiveMember, Complain), as.factor))
```
# Detail Analysis

The dataset contains customer demographic, financial, and behavioral information used to analyze churn patterns.
**Total customers** ≈ 10,000
**Countries:** 

* France
* Germany
* Spain


## Quick Summary


```{r, echo=FALSE}
summary_tbl <- bank_churn_cleaned %>%
  summarise(
    Total_Customers  = n(),
    Total_Exited     = sum(Exited == 1),
    Total_Active     = sum(IsActiveMember == 1),
    Total_Complaints = sum(Complain == 1),
    # Adding your churn rate logic here:
    Churn_Rate       = mean(as.numeric(as.character(Exited)))
  )

# Displaying with formatting
kable(
  summary_tbl %>% 
    mutate(Churn_Rate = scales::percent(Churn_Rate)),
  caption = " ",
  col.names = c("Total Customers", "Total Exited", "Active Members", "Total Complaints", "Churn Rate")
)
```


Approximately **20% of customers have churned**, indicating moderate customer attrition risk.



### Tenure Analysis


```{r, echo=FALSE}
bank_churn_cleaned %>%
  group_by(Tenure) %>%
  summarise(Exited = sum(as.numeric(as.character(Exited)))) %>%
  
  ggplot(aes(x = Tenure, y = Exited)) +
  geom_line(color = "steelblue", size = 1) +
  geom_point(color = "steelblue", size = 2) +
  # Fixes both axes: X shows every year, Y starts at 0
  scale_x_continuous(breaks = 0:10) + 
  scale_y_continuous(limits = c(0, NA)) + 
  labs(
    title = " ", 
    subtitle = " ",
    x = "Tenure (Years)", 
    y = "Number of Exits"
  ) +
  theme_minimal()
```


The data contradicts the standard **loyalty curve** where churn should decrease as tenure increases. To improve retention, the focus must shift from just winning back new joiners to addressing the **sustained dissatisfaction** that causes veteran customers of 5+ years to leave at the same rate as those of 0-2 years.


### Churn by Gender


```{r, include=FALSE}
gender_analysis <- bank_churn_cleaned %>%
  group_by(Gender) %>%
  summarise(Churn_Rate_gender = mean(as.numeric(as.character(Exited))))
```
```{r, echo= FALSE}
ggplot(gender_analysis, aes(x = reorder(Gender, -Churn_Rate_gender), y = Churn_Rate_gender, fill = Gender)) +
  geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate_gender * 100, 1), "%")), 
            vjust = -0.5) + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) + 
  labs(
    title = "  ",
    x = "Gender",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    panel.grid.major.y = element_blank()
  )
```


There is a stark contrast in retention across gender lines: Female customers exhibit a 25.1% churn rate, while Male customers churn at 16.5%. This represents an **8.6% absolute difference**, meaning women are roughly **1.5 times** more likely to leave the bank than men.


### Complain Impact On Churn


```{r, echo=FALSE}
bank_churn_cleaned %>%
  # Filter only for people who actually complained
  filter(Complain == 1) %>% 
  summarise(Total_Exits = sum(as.numeric(as.character(Exited)))) %>%
  ggplot(aes(x = "Complained & Exited", y = Total_Exits)) +
  geom_col(fill = "firebrick", width = 0.5) +
  # This adds the 2034 number on top
  geom_text(aes(label = Total_Exits), vjust = -0.5, size = 6, fontface = "bold") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  labs(
    title = "  ",
    x = NULL,
    y = "Number of Customers"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank(), panel.grid = element_blank())
```


**2,034** customers transitioned from Complaint Filed to Account Closed, representing a **100% failure** rate in complaint-driven retention


### Product Vs Churn


```{r}
bank_churn_cleaned %>%
  group_by(NumOfProducts) %>%
  summarise(Churn_Rate = mean(as.numeric(as.character(Exited)))) %>%
  ggplot(aes(x = factor(NumOfProducts), y = Churn_Rate, fill = factor(NumOfProducts))) +
  geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate * 100, 1), "%")), 
            vjust = -0.5, size = 4.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  # Using a professional color palette
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "  ",
    x = "Number of Products",
    y = "Churn Rate (%)"
  ) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.y = element_blank())
```


The data reveals a *U-Shaped risk curve*. Customers with 2 products have the highest loyalty (lowest churn). However, churn increases significantly for Single-Product customers and **spikes dangerously** for those with 3 or 4 products (often exceeding (80-90%)


### Churn by Geography


```{r, include=FALSE}
geographic_analysis <- bank_churn_cleaned %>%
  group_by(Geography) %>%
  summarise(Churn_Rate_geography = mean(as.numeric(as.character(Exited))))

```
```{r, echo=FALSE}
ggplot(geographic_analysis, aes(x = reorder(Geography, -Churn_Rate_geography), y = Churn_Rate_geography, fill = Geography)) +
  geom_col() +
  geom_text(aes(label = paste0(round(Churn_Rate_geography * 100, 1), "%")), 
            vjust = -0.5) + 
  # This 0.2 multiplier ensures the label doesn't "shift" off the top
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) + 
  labs(
    title = "  ",
    x = "Geography",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.y = element_blank(),
    panel.grid.major.y = element_blank()
  )
```


**Germany** requires an immediate **Deep Dive Audit**. We need to look at the German segment's demographics, average account balances, and product holdings to see if we are losing our most profitable customers or if this is a general exodus



## Deep Dive Analysis | Germany


Analysis On Why is Germany’s Churn Rate Significantly Higher ? 


### Does Account Balance Influence Churn in Germany?


```{r, echo=FALSE}
bank_churn_cleaned %>%
  filter(Geography == "Germany") %>%
  mutate(Balance_Level = ifelse(Balance > mean(Balance), "High Balance", "Low Balance")) %>%
  group_by(NumOfProducts, Balance_Level) %>%
  summarise(Churn_Rate = mean(as.numeric(as.character(Exited)))) %>%
  ggplot(aes(x = factor(NumOfProducts), y = Churn_Rate, fill = Balance_Level)) +
  geom_col(position = "dodge") +
  # We use round(..., 0) as you did, but ensure position_dodge matches
  geom_text(aes(label = paste0(round(Churn_Rate*100, 0), "%")), 
            position = position_dodge(width = 0.9), 
            vjust = -0.5, 
            fontface = "bold") +
  # THIS LINE FIXES THE CUT-OFF ISSUES
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) + 
  labs(
    title = "  ",
    subtitle = "  ",
    x = "Number of Products Held",
    y = "Churn Rate (%)",
    fill = "Customer Wealth"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank()) 
```


Notice that the High Balance and Low Balance are almost the same height in every category. This means German customers aren't leaving because they are rich or poor; they are leaving based on how many products they hold


### Can Customer Engagement Help Reduce Churn in Germany?


```{r, echo=FALSE}
bank_churn_cleaned %>%
  filter(Geography == "Germany") %>%
  group_by(IsActiveMember) %>%
  summarise(Churn_Rate = mean(as.numeric(as.character(Exited)))) %>%
  ggplot(aes(x = factor(IsActiveMember, labels = c("Inactive", "Active")), y = Churn_Rate, fill = factor(IsActiveMember))) +
  geom_col() +
  # We use round(..., 1) for the label
  geom_text(aes(label = paste0(round(Churn_Rate * 100, 1), "%")), 
            vjust = -0.5, size = 5, fontface = "bold") +
  # This line creates the extra space at the top so the text isn't cut off
  scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
  scale_fill_manual(values = c("gray70", "firebrick")) +
  labs(
    title = "  ", 
    x = "Member Status", 
    y = "Churn Rate (%)"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_blank()) # Removes the y-axis numbers since we have labels on the bars
```


Activity is a major factor in retention, but it is **not a cure**. Inactive members in Germany are in a state of collapse with a 41.1% churn rate, while Active members still churn at 23.7%


### How Credit Quality Affects Customer Churn in Germany?


```{r, echo=FALSE}
bank_churn_cleaned %>%
  filter(Geography == "Germany") %>%
  mutate(Credit_Tier = cut(CreditScore, breaks = c(0, 580, 670, 740, 800, 850), 
                           labels = c("Poor", "Fair", "Good", "Very Good", "Exceptional"))) %>%
  group_by(Credit_Tier) %>%
  summarise(Churn_Rate = mean(as.numeric(as.character(Exited)))) %>%
  ggplot(aes(x = Credit_Tier, y = Churn_Rate, group = 1)) +
  geom_line(color = "darkred", size = 1) +
  geom_point(size = 3) +
  labs(title = "  ", x = "Credit Tier", y = "Churn Rate (%)") +
  theme_minimal()

```


Highest churn is among *Poor* credit holders **(~36%)?**, there is a significant and concerning spike in the *Very Good tier* **(~33.5%)**



# Strategic Conclusion


The data confirms that the bank's issue in Germany is not a lack of customer engagement, but a **gap in premium relationship management**. We are successfully attracting high-value, active, and credit-worthy customers, but our current ecosystem—specifically our complaint resolution and multi-product bundling—is creating *friction* that drives them to competitors. By stabilizing the German market through a *Plus-One* product strategy and a dedicated complaint-recovery desk, we can protect the bank's most profitable assets and reduce regional churn back to sustainable levels.

# About the Author


I am **Maria Aslam**, a Physics graduate with a strong interest in data analytics. I hold the **Google Data Analytics Professional Certificate** and have learned core analytical tools including R, Excel, SQL, Tableau, and Power BI through coursework and project-based learning. This project demonstrates my approach to applying analytical methods to business data as part of my ongoing skill development.


###### **Connect on LinkedIn:** 

[LinkedIn](https://www.linkedin.com/in/maria-aslam-458860316/){target="_blank"}


###### **Explore More Dashboards:** 

[Tableau Public Profile](https://public.tableau.com/app/profile/maria.aslam/vizzes){target="_blank"}


###### **View the Source Code:**

[GitHub Repository](https://github.com/Maria1681){target="_blank"}

